---
title: "ESA Listings"
output:
  flexdashboard::flex_dashboard:
    css: custom.css
    social: menu
    df_print: paged
    navbar:
    - {title: "", icon: "fa-question-circle fa-lg", align: right, href: "mailto:esa@defenders.org?subject=listings summary"}
    - {title: "", icon: "fa-github fa-lg", align: right, href: "https://github.com/jacob-ogre/ESA_listings_v2"}
    - {title: "CCI", align: right, href: "https://cci-dev.org"}
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(dplyr)
library(DT)
library(ecosscraper)
library(ESAListings)
library(highcharter)
library(leaflet)
library(plotly)
library(RPostgreSQL)
library(secret)
library(shinyBS)
library(shinydashboard)
library(treemap)
library(viridis)

con <- dbConnect(
  dbDriver("PostgreSQL"),
  dbname = "postgres",
  user = ESC_DB_USR,
  password = get_secret(ESC_DB_PW, key = priv_key, vault),
  host = ESC_DB
)

TECP_table <- dbSendQuery(con, "select * from tecp_table") %>% dbFetch()
TECP_data <- dplyr::filter(TECP_table, u_s_or_foreign_listed != "Foreign")

col_names <- c("Scientific_Name", "Common_Name", "Species_Code",  
               "Critical_Habitat", "First_Listed", "Species_Group",
               "Lead_Region", "Federal_Listing_Status", "Special_Rules",
               "US_or_Foreign", "Where_Listed", "Species_Page")
names(TECP_data) <- col_names

load("data/county_topo.rda")
load("data/esacounties.rda")
# load("data/TECP_data.rda")
# load("data/TECP_domestic.rda")
load("data/county_attrib.rda")

counties <- readRDS("data/counties_df.rds")

# a palette of c("yellow","red","black","green","purple","orange")
list_pal <- rev(substr(viridis(4), 1, 7)) 

#define pallete funciton converting status names to colors
stat_pal <- function(status) {
  switch(
    status,
    Candidate = list_pal[2],
    Endangered = list_pal[1],
    Proposed = list_pal[3],
    Threatened = list_pal[4]
  )
}

#define palette function for choropleth map
palfx <- colorNumeric(
  palette = c("midnightblue", "yellow"),
  domain = c(0, 75),
  na.color = "yellow"
)
```

```{r en_th_can_prop, echo=FALSE}
# This needs to run each app start-up because the TECP data should scrape nightly
num_es <- nrow(filter(TECP_data, Federal_Listing_Status == "Endangered"))
num_th <- nrow(filter(TECP_data, Federal_Listing_Status == "Threatened"))
num_pr <- nrow(
  filter(
    TECP_data,
    startsWith(Federal_Listing_Status, "Proposed")
  )
)
num_cn <- nrow(filter(TECP_data, Federal_Listing_Status == "Candidate"))

##############################################################################
# create 'regions' dataframe
regions <- group_by(
  TECP_data,
  Lead_Region,
  Species_Group,
  Federal_Listing_Status
)

regions$Group <- sapply(regions$Species_Group, function(x)
  if (x == "Ferns and Allies" |
      x == "Flowering Plants" | 
      x == "Conifers and Cycads" |
      x == "Lichens") {
    "Plants and Lichens"
  } else if (x == "Snails" | x == "Clams") {
    "Molluscs"
  } else{
    x
  })

regions$Status <- sapply(regions$Federal_Listing_Status, function(x)
  if (x == "Proposed Endangered" | x == "Proposed Threatened") {
    "Proposed"
  } else{
    x
  })

regions <- group_by(regions, Lead_Region, Group, Status) %>%
  filter(Status != "Experimental Population, Non-Essential" &
         Status != "Similarity of Appearance to a Threatened Taxon") %>%
  summarise(count = n())

regions <- as.data.frame(regions)
regions$Lead_Region[regions$Lead_Region != "NMFS"] <-
  paste("Region", regions$Lead_Region[regions$Lead_Region != "NMFS"])

##############################################################################
# create 'years' dataframe
years <- mutate(TECP_data, Year = substr(First_Listed, 9, 12)) %>%
  select(Year, Federal_Listing_Status) %>%
  filter(Federal_Listing_Status == "Endangered" |
         Federal_Listing_Status == "Threatened")

years$Status <- sapply(years$Federal_Listing_Status,
                       function(x) if (x == "Proposed Endangered" |
                                       x == "Proposed Threatened") {
                         "Proposed" } else { x } )

years <- group_by(years, Year, Status) %>%
  summarise(count = n())

years$Year <- as.integer(years$Year)

impute <- data.frame(Year = rep(seq(
  min(years$Year, na.rm = TRUE),
  max(years$Year, na.rm = TRUE), 1), 2),
Status = rep(unique(years$Status),
             each = max(years$Year, na.rm = TRUE) - 1966))

years <- right_join(years, impute, by = c("Year", "Status"))
years$count[is.na(years$count)] <- 0

totals <- group_by(years, Year) %>%
  arrange(Year) %>%
  summarise(total = sum(count)) %>%
  mutate(cumm = cumsum(total))

#create species dataset
species <- group_by(esacounties, Scientific) %>%
  summarise(count = n()) %>%
  arrange(count)

#create initial treemaps
dat1 <- group_by(regions, Lead_Region, Status) %>%
  summarize(count = sum(count))
dat3 <- group_by(regions, Group, Status) %>%
  summarize(count = sum(count))

tm_tx <- list()
for (i in unique(dat1$Lead_Region)) {
  ls1 <-
    list(
      name = i,
      id = i,
      value = sum(dat1$count[dat1$Lead_Region == i]),
      color = NA
    )
  tm_tx[[length(tm_tx) + 1]] <- ls1
}
for (i in 1:length(dat1$count)) {
  ls2 <-
    list(
      parent = dat1$Lead_Region[i],
      name = dat1$Status[i],
      value = dat1$count[i],
      color = stat_pal(strsplit(dat1$Status[i], " ")[[1]][1])
    )
  tm_tx[[length(tm_tx) + 1]] <- ls2
}

tm_rg <- list()
for (i in unique(dat3$Group)) {
  ls1 <- list(
      name = i,
      id = i,
      value = sum(dat3$count[dat3$Group == i]),
      color = NA
    )
  tm_rg[[length(tm_rg) + 1]] <- ls1
}

for(i in 1:length(dat3$count)) {
  ls2 <- list(
    parent = dat3$Group[i],
    name = dat3$Status[i],
    value = dat3$count[i],
    color = stat_pal(strsplit(dat3$Status[i], " ")[[1]][1])
  )
  tm_rg[[length(tm_rg) + 1]] <- ls2
}

rg_combos <- data.frame(
  Lead_Region = rep(unique(regions$Lead_Region), each = 4),
  Status = rep(c("Endangered", "Proposed", "Threatened", "Candidate"), 9),
  stringsAsFactors = FALSE
)

tx_combos <- data.frame(
  Group = rep(unique(regions$Group), each = 4), 
  Status = rep(c("Endangered", "Proposed", "Threatened", "Candidate"), 11),
  stringsAsFactors = FALSE
)
```

Overview
=======================================================================

Column {data-width=200}
-----------------------------------------------------------------------

### By the numbers

```{r numbers_summary, echo=FALSE}
fluidRow(
  valueBox(num_es, "Endangered Species", color = "yellow", width = 11),
  valueBox(num_th, "Threatened Species", color = "purple", width = 11),
  valueBox(num_cn, "Candidate Species", color = "green", width = 11),
  valueBox(num_pr, "Proposed for Listing", color = "blue", width = 11)
)
```

----

<p style='font-size:0.92em'>Since 1973, over 2,000 species have been listed as 'Threatened' or 'Endangered' under the Endangered Species Act (ESA). The Fish and Wildlife Service (FWS) provides a detailed description of the listing process. Here you can explore the number of domestic listed species by year as of `r Sys.Date()`.</p>

Column {data-width=800}
-----------------------------------------------------------------------

### Domestic ESA listings through time

```{r listings_time, echo=FALSE}
renderPlotly({
  plot_ly(ungroup(years), x = ~Year, y = ~count) %>%
    add_trace(
      type = "scatter", 
      mode = "lines", 
      color = ~Status, 
      colors = list_pal, 
      text = ~paste(count, "species listed as<br>", Status, "in", Year, sep = " "), 
      hoverinfo = "text") %>%
    add_trace(
      data = totals, 
      x = ~Year, 
      y = ~total, 
      text = ~paste(total,"Total species listed in", Year, sep = " "), 
      hoverinfo = "text",
      type = "scatter", 
      mode = "lines", 
      name = "Total", 
      line = list(color = "grey")) %>%
    add_trace(
      data = totals, 
      x = ~Year, 
      y = ~cumm, 
      text = ~paste(cumm, "species listed as of", Year, sep = " "), 
      hoverinfo = "text",
      type = "scatter", 
      mode = "lines", 
      name = "Cumulative<br>(click to show)", 
      visible = "legendonly") %>%
    layout(
      hovermode = "closest", 
      font = list(color = "black"),
      xaxis = list(title = "Year"),
      yaxis = list(title = "Number of Listings"),
      legend = list(x = 0.05, y = 0.95, bordercolor = "black", borderwidth = 1))
})
```

By Group & Region
=======================================================================

Column {.tabset}
-----------------------------------------------------------------------

### Taxonomic focus

```{r bar_by_taxon_status}

dat2 <- reactive(
  filter(regions, grepl(input$rg_select, Lead_Region)) %>%
    group_by(Group, Status) %>%
    summarize(count = sum(count))
)

fluidRow(
  column(6,
    p("How many species of each taxonomic group are listed as endangered,
      threatened, proposed, or candidate? (You may 'turn off' plants and
      lichens - a big group that obscures the counts for other groups - with 
      the 'Data selector' buttons.)")
  ),
  column(6,
    p("Treemaps organize information hierarchically; here we use it 
      to visualize the breakdown of taxa within specific FWS 
      Regional Offices. Box sizes correspond to the number of listed 
      species. Mouse over the boxes to see the number of listed species, and 
      click on a taxon to zoom in.")
  )
)
    
fluidRow(
  column(6,
    renderHighchart({
      spec_in_data <- left_join(tx_combos[tx_combos$Group!= input$plants, ], dat2())
      hchart(spec_in_data,
             type = "column",
             hcaes(x = Group, y = count, group = Status)) %>%
        hc_yAxis(title = list(text = "Number of Listed Species",
                              style = list(color = "black")),
                 labels = list(style = list(color = "black")),
                 stackLabels = list(enabled = "true")) %>%
        hc_xAxis(title = list(text = NULL),
                 labels = list(style = list(color = "black"))) %>%
        hc_plotOptions(column = list(stacking = "normal")) %>%
        hc_tooltip(
          headerFormat = "<b>{point.x}</b><br>",
          pointFormat = "{series.name}: {point.y}<br>Total: {point.stackTotal}") %>%
        hc_colors(list_pal[c(2,1,3,4)])
    })
  ),
  column(6,
    renderHighchart({
      # make the data.frame in reactive context 
      tm_rg <- list()
      for(i in unique(dat2()$Group)){
        ls1 <- list(name = i,
                    id = i,
                    value = sum(dat2()$count[dat2()$Group == i]),
                    color = NA)
        tm_rg[[length(tm_rg) + 1]] <- ls1
      }
      for(i in 1:length(dat2()$count)){
        ls2 <- list(parent = dat2()$Group[i],
                    name = dat2()$Status[i],
                    value = dat2()$count[i],
                    color = stat_pal(strsplit(dat2()$Status[i], " ")[[1]][1]))
        tm_rg[[length(tm_rg) + 1]] <- ls2
      }
      
      highchart() %>%
        hc_add_series(
          data = tm_rg,
          type = "treemap",
          allowDrillToNode = TRUE,
          layoutAlgorithm = "squarified",
          levels = list(
            list(
              level = 1,
              borderColor = "white",
              borderWidth = 5,
              dataLabels = list(
                enabled = "true",
                align = "left",
                verticalAlign = "top",
                style = list(fontSize = "14px")
              )
            ),
            list(
              level = 2,
              borderColor = "grey",
              borderWidth = 3,
              dataLabels = list(
                enabled = FALSE,
                align = "center",
                verticalAlign = "middle")
              )
            )
          ) %>%
        hc_tooltip(pointFormat = "<b>{point.name}<\b><br>
                   Listings: {point.value}")
    })
  )
)
```

```{r taxa_data_sel, echo=FALSE}
br()
hr()
div(style="border: 1px solid #F5F5F5; border-radius:3px; background: #F5F5F5",
fluidRow(
  column(4,
    h3(style="padding-left:10px;", "Data selectors")
  ),
  column(5,
    h4("Bar chart: Plants and Lichens"),
    radioButtons(
      "plants",
      label = NULL,
      choices = c("Show" = "", "Hide" = "Plants and Lichens"),
      inline = TRUE
    )
  ),
  column(2,
    h4("Bar chart + Treemap"),
    selectInput(
      inputId = "rg_select",
      NULL,
      choices = setNames(
        as.list(c("", unique(regions$Lead_Region))),
                c("All FWS Regions", 
                  unique(regions$Lead_Region)[1:8],
                  "NMFS (Lead)")
      ),
      selected = "",
      selectize = FALSE
    )
  )
))

```

### Regional focus

```{r bar_by_region}

dat1 <- reactive(
  filter(regions, grepl(input$tx_select, Group)) %>%
    group_by(Lead_Region, Status) %>%
    summarize(count = sum(count))
)

fluidRow(
  column(6,
    p("How many species are listed as endangered, threatened, proposed, 
      or candidate in each FWS region of the US?")
  ),
  column(6,
    p("Treemaps organize information hierarchically; here we use it 
      to visualize the breakdown of listing categories by FWS 
      Regional Office. Box sizes correspond to the number of listed 
      species. Mouse over the boxes to see the number of listed species, and 
      click on a region to zoom in.")
  )
)

fluidRow(
  column(6,
    renderHighchart({
      reg_in_data <- left_join(rg_combos, dat1())
      hchart(
        reg_in_data,
        type = "column",
        hcaes(x = Lead_Region,
              y = count,
              group = Status))  %>%
        hc_yAxis(title = list(
          text = "Number of Listed Species",
          style = list(color = "black")
        ),
        labels = list(style = list(color = "black")),
        stackLabels = list(enabled = "true")) %>%
        hc_xAxis(categories = c("Reg. 1", "Reg. 2", "Reg.3", "Reg. 4",
                                "Reg. 5", "Reg. 6", "Reg. 7", "Reg. 8", "NMFS"),
                 title = list(text = NULL),
                 labels = list(style = list(color = "black"))) %>%
        hc_plotOptions(series = list(stacking = "normal")) %>%
        hc_tooltip(
          headerFormat = "<b>{point.x}</b><br>",
          pointFormat = "{series.name}: {point.y}<br>Total: {point.stackTotal}") %>%
        hc_colors(list_pal[c(2,1,3,4)])
    })
    
  ),
  column(6,
    renderHighchart({
      # make the data.frame in reactive context 
      tm_tx <- list()
      for(i in unique(dat1()$Lead_Region)){
        ls1 <- list(name = i,
                    id = i,
                    value = sum(dat1()$count[dat1()$Lead_Region == i]),
                    color = NA)
        tm_tx[[length(tm_tx) + 1]] <- ls1
      }
      for(i in 1:length(dat1()$count)){
        ls2 <- list(parent = dat1()$Lead_Region[i],
                    name = dat1()$Status[i],
                    value = dat1()$count[i],
                    color = stat_pal(strsplit(dat1()$Status[i], " ")[[1]][1]))
        tm_tx[[length(tm_tx) + 1]] <- ls2
      }
      
      highchart() %>%
        hc_add_series(
          data = tm_tx,
          type = "treemap",
          allowDrillToNode = TRUE,
          layoutAlgorithm = "squarified",
          levels = list(
            list(
              level = 1,
              borderColor = "white",
              borderWidth = 5,
              dataLabels = list(
                enabled = "true",
                align = "left",
                verticalAlign = "top",
                style = list(fontSize = "14px")
              )
            ),
            list(
              level = 2,
              borderColor = "grey",
              borderWidth = 3,
              dataLabels = list(
                enabled = FALSE,
                align = "center",
                verticalAlign = "middle")
              )
            )
          ) %>%
        hc_tooltip(pointFormat = "<b>{point.name}<\b><br>
                   Listings: {point.value}")
    })
  )
)
```

```{r region_data_sel, echo=FALSE}
br()
hr()
div(style="border: 1px solid #F5F5F5; border-radius:3px; background: #F5F5F5",
fluidRow(
  column(9,
    h3(style="padding-left:10px;", "Data selectors")
  ),
  column(2,
    h4("Bar chart + Treemap"),
    selectInput(
      "tx_select",
      NULL,
      choices = setNames(
        as.list(c("", unique(regions$Group))),
                c("All Taxa", unique(regions$Group))),
      selected = "",
      selectize = FALSE
    )
  )
))

```

U.S. Listings Map
=======================================================================

Column {data-width=750}
-----------------------------------------------------------------------

### 

```{r listing_map}
leafletOutput("map")

output$map <- renderLeaflet({
  leaflet() %>%
    addTopoJSON(
      county_topo,
      weight = 0.5,
      color = "black",
      fillOpacity = 0.5) %>%
    addProviderTiles("Stamen.TonerLite") %>%
    setView(lng=-95, lat=38, zoom = 4)  %>%
    mapOptions(zoomToLimits = "never") %>%
    addLegend(
      title = "Number of<br>Listed Species",
      position = "bottomleft",
      colors = palfx(seq(0, 80, 10)),
      labels = c("- 0","- 10","- 20","- 30","- 40","- 50",
                 "- 60","- 70","> 80")) %>%
    addLegend(
      title = "Click on the center of a county<br>to see the species listed there.",
      position = "topright", colors = NULL, labels = NULL) %>%
    addCircleMarkers(
      data = counties,
      lng = ~INTPTLON,
      lat = ~INTPTLAT,
      radius = 6,
      color = "red",
      fillOpacity = 0,
      stroke = FALSE,
      popup = ~paste0(NAME," County<br>", count, " species<br>")
    )
})

# cur_zoom <- reactive({
#   if (!is.null(input$map_zoom)) {
#     input$map_zoom
#   } else {
#     4
#   }
# })

observe({
  if(!is.null(input$cnty_spp_cell_clicked$value)) {
    print(1 / input$map_zoom)
    spec_select <- filter(esacounties,
                          Scientific == input$cnty_spp_cell_clicked$value) %>%
                   select(GEOID)
  
    leafletProxy("map") %>%
      clearShapes() %>%
      clearMarkers() %>%
      addCircles(
        data = filter(counties, GEOID %in% spec_select$GEOID),
        lng = ~INTPTLON,
        lat = ~INTPTLAT,
        radius = 32000 * (2 / input$map_zoom),
        color = "black",
        fillOpacity = 1,
        stroke = FALSE) %>%
      addCircleMarkers(
        data = counties,
        lng = ~INTPTLON,
        lat = ~INTPTLAT,
        radius = 6,
        color = "red",
        fillOpacity = 0,
        stroke = FALSE,
        popup = ~paste0(NAME," County<br>", count, " species<br>")
      )
  }
})

observeEvent(input$map_marker_click,{
  click_lat <- input$map_marker_click[[3]]
  click_lon <- input$map_marker_click[[4]]
  gid <- counties$GEOID[counties$INTPTLAT == click_lat & 
                          counties$INTPTLON == click_lon]
  
  output$cnty_spp <- DT::renderDataTable({
    filter(esacounties, GEOID == gid) %>%
    select(Scientific, Common) %>%
    datatable(
      rownames = FALSE, 
      selection = "single", 
      colnames = c("Species", "Common Name"), 
      caption = "Click a species name to see its range displayed")
  })
})

```

Column {data-width=250}
-----------------------------------------------------------------------

### 

```{r spp_table, echo=FALSE}
dataTableOutput("cnty_spp")
```

Range Sizes
=======================================================================

Column {data-width=750}
-----------------------------------------------------------------------

### Counties per species and species per county

```{r}
plotlyOutput("cumulants")

output$cumulants <- renderPlotly({
  #cumulative number of counties per species
  co_spp <- plot_ly(
    arrange(species, count), 
    x = ~count, 
    source = "sp_dist") %>%
    add_lines(
      y = ~round(row_number(count)/nrow(species), 3), 
      name = "counties / spp.",
      type = "scatter", 
      mode = "lines",
      text = ~paste(
        round(row_number(count)/nrow(species), 3) * 100,
        "% of species occur in", 
        count, 
        "or fewer counties"), 
      hoverinfo = "text") %>%
    add_histogram(
      histnorm = "probability", 
      name = "# counties", 
      xbins = list(
        start = 0.5, 
        end = 3102.5, 
        size = 1),
      text = ~paste(
        "% of counties contain", 
        "listed species"),
      hoverinfo = "none") %>%
    layout(
      xaxis = list(
        title = "# Counties in Species' Range", 
        type = "log", 
        tickvals = c(0, 1, 5, 10, 50, 100, 500, 1000, 3000)),
       yaxis = list(title = "Percentile of Species")
    )

  # cummulative number of species per county
  spp_co <- plot_ly(
      arrange(counties, count), 
      x = ~count) %>%
    add_lines(
      y = ~round(row_number(count)/nrow(counties), 3), 
      name = "spp. / county",
      type = "scatter", 
      mode = "lines",
      text = ~paste(
        round(row_number(count)/nrow(counties), 3) * 100, 
        "% of counties contain", 
        count, 
        "or fewer species"),
      hoverinfo = "text") %>%
    add_histogram(
      histnorm = "probability", 
      name = "# spp.",
      hoverinfo = "none") %>%
    layout(
      xaxis = list(
        title = "# Species per County", 
        type = "log", 
        tickvals = c(0, 1, 5, 10, 50, 100, 1000)),
      yaxis = list(title = "Percentile of Counties"),
      margin = list(t = 10),
      legend = list(x = 0.02, y = 0.98))
  
  subplot(
    nrows = 2, 
    # heights = c(0.5, 0.45),
    margin = 0.05,
    titleX = TRUE,
    co_spp, 
    spp_co)
})

```

Column {data-width=250}
-----------------------------------------------------------------------

### Distributions of Listed Species

The distribution of ESA-listed species is very skewed toward small ranges (shown in terms of the number of U.S. counties of occurrence; top graph). 41% of species occur in one county, 55% occur in one or two counties, and 95% of species occur in less than 50 of the >3,000 counties in the U.S. 

<br><br><br><br>

Similarly, most counties have few ESA-listed species, although this distribution is less skewed than the sizes of species' ranges (bottom graph). From the data we see that 25% of counties are home to four or fewer ESA-listed species and 75% of counties are home to eight or fewer ESA-listed species.

```{r echo=FALSE}
# dataTableOutput("rngTble")
# 
# observeEvent(event_data("plotly_click", source = "sp_dist"), {
#   sp_count <- event_data("plotly_click", source = "sp_dist")$x[[1]]
# 
#   output$rngTble <- DT::renderDataTable({
#     filter(ungroup(species), count == sp_count) %>%
#       inner_join(TECP_data, by = c("Scientific" = "Scientific_Name")) %>%
#       select(Scientific, Common_Name, count) %>%
#       datatable(rownames = FALSE,
#                 selection = "none",
#                 colnames = c("Species", "Common Name", "Counties"))
#   })
# })
```
